Rotation Period Evolution in Low-Mass Binary Stars: The Impact of Tidal Torques and Magnetic Braking

### 2018-12-28 Paper/sync.tex

The long-term angular momentum evolution of low-mass (M$\lsim 1$ M$_{\odot}$) stars is controlled by magnetic braking, the torque exerted on stars due to the coupling of stellar winds to the surface magnetic field \citep{Dunn1961,Mestel1968}. Early on in stellar lifetimes, stars spin-up as they contract along the premain sequence.  Once stars reach the main sequence, stellar radii remain mostly constant and magnetic braking saps angular momentum from low-mass stars, gradually spinning them down over time \citep{Skumanich1972}. Although the precise details of how magnetic braking operates are not fully known, models of magnetic braking have been used to successfully model the bulk trends of P$_{rot}$ distributions in clusters \citep[e.g. Praesepe, ][]{Matt2015,Douglas2017} and field stars \citep[e.g. the \kepler field, ][]{Reiners2012,Matt2015,vanSaders2018}. Furthermore, the magnetic braking-driven long-term spin-down of stars has been used to estimate stellar ages, a method known as gyrochronlogy \citep{Skumanich1972,Barnes2003,Barnes2007,Mamajek2008,Barnes2010}, with older stars assumed to have lost more angular momentum due to magnetic braking and therefore rotate more slowly.

@@

Tides impart a significant signature in the long-term angular momentum evolution for binary stars, especially for stellar spins. Tidal torques drive binaries towards the tidally-locked state in which the stellar P$_{rot}$ is equal to the equilibrium rotation period (P$_{eq}$) predicted by tidal models, with a familiar example of this effect being spin-orbit synchronization where P$_{rot} = $ P$_{eq} = $ P$_{orb}$.  Tidal-locking occurs much earlier than orbital circularization as the tidal-locking timescale is estimated to be $2-3$ orders of magnitude less than the circularization timescale \citep{Zahn1989,Witte2002,Mazeh2008} as there is less angular momentum in stellar spins than the binary orbit. As a result, tidal-locking is expected for binaries with P$_{orb} \lsim$ 20 d \citep[e.g.][]{Levato1974,Meibom2006,Mazeh2008,Zahn2008,Meibom2015}.

@@

We simulate coupled stellar-tidal evolution for low-mass binaries using an improved version of the coupled stellar-tidal evolution model presented in \citet{Fleming2018}.  We implement our model in the open-source code VPLanet\footnote{VPLanet is publicly available
at \href{https://github.com/VirtualPlanetaryLaboratory/vplanet}{{https://github.com/VirtualPlanetaryLaboratory/vplanet}}.} \citep[][Barnes et al., in prep]{Barnes2016,vplanet2018}.  We integrate all model equations (see $\S$~\ref{sec:methods:stellar} and $\S$~\ref{sec:methods:eqtide}) using using the $4^{th}$ order Runge-Kutta scheme with adaptive timestepping described in \citet{Fleming2018}.

@@

\begin{figure*}
	\includegraphics[width=\textwidth]{../Plots/eqPerTwoPanel.pdf}
   \caption{Evolution of stellar P$_{rot}$, normalized by P$_{eq}$ (see Eqn.~(\ref{eqn:cpl:eqPer}) and Eqn.~(\ref{eqn:ctl:eqPer}), for initial circular binary orbits according to the CPL (solid) and CTL (dashed) models with $Q = 10^6$ and $\tau = 0.1$ s, respectively. Left: P$_{rot}/$P$_{eq}$ for stars with P$_{orb}$ ranging from 5 d to 60 d. The black dotted line indicates the tidally-locked state where P$_{rot} = $ P$_{eq}$.  Right: Net P$_{rot}$ derivative due to stellar evolution, tidal torques, and magnetic braking.  We truncate each curve when the binary tidally locks. For P$_{orb} \leq 10$ d, both the CPL and CTL models predict that the binaries lock into synchronous rotation.  For all P$_{orb}$, the CPL models tidally-lock into synchronous rotation whereas the CTL model predicts subsynchronous (P$_{rot} > $ P$_{eq}$) rotation that persists for Gyrs.}%
    \label{fig:eqPer}%
\end{figure*}

@@

\begin{figure}
	\includegraphics[width=0.45\textwidth]{../Plots/eqPerShortPorb.pdf}
   \caption{Evolution of stellar P$_{rot}$, normalized by P$_{eq}$ (see Eqn.~(\ref{eqn:cpl:eqPer}) and Eqn.~(\ref{eqn:ctl:eqPer}), for initial circular binary orbits with initial P$_{orb} = 5$ d according to the CPL (solid) and CTL (dashed) models for several values of $Q$ and $\tau$, respectively. Systems with strong tidal torques ($Q = 10^6$ or $\tau = 0.1$ s) tidally lock, whereas in systems with weaker tidal torques (larger $Q$ and smaller $\tau$, respectively), magnetic braking initially overpowers tidal torques, spinning down the stars past the tidally-locked state, resulting in subsynchronous rotation. }%
    \label{fig:eqPerShortPorb}%
\end{figure}

@@

\begin{figure}
	\includegraphics[width=0.45\textwidth]{../Plots/porbProtPorbQHist.pdf}
   \caption{Median log$_{10}(Q)$ of primary stars binned by P$_{orb}$ and P$_{orb}/$P$_{rot}$.  In each bin, we compute the median, $16\%$, and $84\%$ percentiles for log$_{10}(Q)$ and display these values. Narrow constraints imply that certain tidal parameters are required to produce that rotation state, e.g. supersynchronous rotators near P$_{orb}/$P$_{rot} = 1.5$ require $Q \lsim 10^{5.5}$. }%
    \label{fig:qmap}%
\end{figure}

>>>>

The long-term angular momentum evolution of low-mass (M$\lsim 1$ M$_{\odot}$) stars is controlled by magnetic braking, the torque exerted on stars due to the coupling of stellar winds to the surface magnetic field \citep{Dunn1961,Mestel1968}. Early on in stellar lifetimes, stars spin-up as they contract along the premain sequence.  Once stars reach the main sequence, stellar radii remain mostly constant while magnetic braking removes angular momentum from the stars, gradually spinning them down over time \citep{Skumanich1972}. Although the precise details of how magnetic braking operates are not fully known, models of magnetic braking have been used to successfully model the bulk trends of P$_{rot}$ distributions in clusters \citep[e.g. Praesepe, ][]{Matt2015,Douglas2017} and field stars \citep[e.g. the \kepler field, ][]{Reiners2012,Matt2015,vanSaders2018}. Furthermore, the magnetic braking-driven long-term spin-down of stars has been used to estimate stellar ages, a method known as gyrochronlogy \citep{Skumanich1972,Barnes2003,Barnes2007,Mamajek2008,Barnes2010}, with older stars assumed to have lost more angular momentum due to magnetic braking and therefore rotate more slowly.

@@

Tides impart a significant signature in the long-term angular momentum evolution for binary stars, especially for stellar spins. Tidal torques drive binaries towards the tidally-locked state in which the stellar P$_{rot}$ is equal to the equilibrium rotation period (P$_{eq}$) predicted by tidal models, with a familiar example of this effect being spin-orbit synchronization where P$_{rot} = $ P$_{eq} = $ P$_{orb}$.  Tidal-locking occurs much earlier than orbital circularization with the tidal-locking timescale estimated to be $2-3$ orders of magnitude less than the circularization timescale \citep{Zahn1989,Witte2002,Mazeh2008} as there is less angular momentum in stellar spins than the binary orbit. As a result, tidal-locking is expected for binaries with P$_{orb} \lsim$ 20 d \citep[e.g.][]{Levato1974,Meibom2006,Mazeh2008,Zahn2008,Meibom2015}.

@@

We simulate coupled stellar-tidal evolution for low-mass binaries using an improved version of the model presented in \citet{Fleming2018}.  We implement our model in the open-source code VPLanet\footnote{VPLanet is publicly available
at \href{https://github.com/VirtualPlanetaryLaboratory/vplanet}{{https://github.com/VirtualPlanetaryLaboratory/vplanet}}.} \citep[][Barnes et al., in prep]{Barnes2016,vplanet2018}.  We integrate all model equations (see $\S$~\ref{sec:methods:stellar} and $\S$~\ref{sec:methods:eqtide}) using using the $4^{th}$ order Runge-Kutta scheme with adaptive timestepping described in \citet{Fleming2018}.

@@

\begin{figure*}
	\includegraphics[width=\textwidth]{../Plots/eqPerTwoPanel.pdf}
   \caption{Evolution of stellar P$_{rot}$, normalized by P$_{eq}$ (see Eqn.~(\ref{eqn:cpl:eqPer}) and Eqn.~(\ref{eqn:ctl:eqPer}), for initial circular binary orbits according to the CPL (solid) and CTL (dashed) models with $Q = 10^6$ and $\tau = 0.1$ s, respectively, using the \citet{Matt2015} magnetic braking model. Left: P$_{rot}/$P$_{eq}$ for stars with P$_{orb}$ ranging from 5 d to 60 d. The black dotted line indicates the tidally-locked state where P$_{rot} = $ P$_{eq}$.  Right: Net P$_{rot}$ derivative due to stellar evolution, tidal torques, and magnetic braking.  We truncate each curve when the binary tidally locks. For P$_{orb} \leq 10$ d, both the CPL and CTL models predict that the binaries lock into synchronous rotation.  For all P$_{orb}$, the CPL models tidally-lock into synchronous rotation whereas the CTL model predicts subsynchronous (P$_{rot} > $ P$_{eq}$) rotation that persists for Gyrs.}%
    \label{fig:eqPer}%
\end{figure*}

@@

\begin{figure}
	\includegraphics[width=0.45\textwidth]{../Plots/eqPerShortPorb.pdf}
   \caption{Evolution of stellar P$_{rot}$, normalized by P$_{eq}$ (see Eqn.~(\ref{eqn:cpl:eqPer}) and Eqn.~(\ref{eqn:ctl:eqPer}), for initial circular binary orbits with initial P$_{orb} = 5$ d according to the CPL (solid) and CTL (dashed) models for several values of $Q$ and $\tau$, respectively, using the \citet{Matt2015} magnetic braking model. Systems with strong tidal torques ($Q = 10^6$ or $\tau = 0.1$ s) tidally lock, whereas in systems with weaker tidal torques (larger $Q$ and smaller $\tau$, respectively), magnetic braking initially overpowers tidal torques, spinning down the stars past the tidally-locked state, resulting in subsynchronous rotation. }%
    \label{fig:eqPerShortPorb}%
\end{figure}

@@

\begin{figure}
	\includegraphics[width=0.45\textwidth]{../Plots/porbProtPorbQHist.pdf}
   \caption{Median log$_{10}(Q)$ of primary stars binned by P$_{orb}$ and P$_{orb}/$P$_{rot}$ evolved using the CPL model.  In each bin, we compute the median, $16\%$, and $84\%$ percentiles for log$_{10}(Q)$ and display these values. Narrow constraints imply that certain tidal parameters are required to produce that rotation state, e.g. supersynchronous rotators near P$_{orb}/$P$_{rot} = 1.5$ require $Q \lsim 10^{5.5}$. }%
    \label{fig:qmap}%
\end{figure}
